# Integrated Population Models {#PopModels}

The field and analytical methods discussed thus far have addressed population rates one at a time (e.g., estimating abundance or survival). An improved approach would be to design a complementary set of field studies that provides a complete picture of population dynamics. For example, @conn.etal_2008 used harvest and tag recovery data to fit an integrated population model\index{integrated population model} (IPM) for black bear (*Ursus americanus*). Their model provided estimates that were more precise and biologically realistic than point estimates using capture-recapture\index{capture-recapture} data alone.  @kéry.schaub_2011 developed an avian IPM using adult population counts, nest surveys, and capture-recapture data on survival.  We begin here with a somewhat simpler approach, a two-age model (ages 1 and 2+) using capture-recapture and telemetry data, then examine a full age-structured stock assessment model (Section \@ref(CAGEAN)).

## Two-age model {#TwoAge}
This example shows how two independent field studies (capture-recapture\index{capture-recapture} estimates of abundance and telemetry\index{telemetry} information on survival) can be combined to make inferences about population dynamics. The cost of this approach in time and money is higher than a basic relative-abundance survey, but it provides insights about population dynamics (recruitment and survival) that would not be possible with relative abundance data alone. The cost could be justified for a subset of cases; for example, a population at low abundance or one supporting an important fishery. We begin with the simulation model:

```{r eval=FALSE}
rm(list=ls()) # Clear Environment

# Simulation to create population size and field data
Y <- 10

LowS <- 0.5 # Adult survival, arbitrary lower bound
HighS <- 0.9 # Arbitrary upper bound
AnnualS <- runif(n=(Y-1), min=LowS, max=HighS)

MeanR <- 400 # Arbitrary mean annual recruitment
N.r <- array(data=NA, dim=Y)
SD_rec <- 0.25 # Low level of lognormal error in recruitment
N.r <- trunc(MeanR * rlnorm(n=Y, 0, SD_rec))
N.a <- array(data=NA, dim=Y)
N.a[1] <- N.r[1] + rpois(n=1, lambda=MeanR) # Arbitrary year-1 population size
for (y in 2:Y){
  N.a[y] <- N.r[y] + rbinom(n=1, size=N.a[y-1], prob=AnnualS[y-1])
} #y

# Short-term (closed) two-sample population estimate
CapProb <- 0.2 # Fraction of population sampled
n1 <- rbinom(n=Y, size=N.a, prob=CapProb) # Caught and marked in first sample
n2 <- rbinom(n=Y, size=N.a, prob=CapProb) 
# Caught and examined for marks in second sample
m2 <- rbinom(n=Y, size=n2, prob=n1/N.a) # Marked fish in second sample

# Telemetry information on annual survival
TelRelease <- 20 # Number of telemetry tags to release annually
TelTags <- array(data=NA, dim=c((Y-1),Y))
for (y in 1:(Y-1)){
  TelTags[y,y] <- TelRelease
  for (t in ((y+1):Y)){
    TelTags[y,t] <- rbinom(n=1, size=TelTags[y,(t-1)], prob=AnnualS[t-1])
  }} # y,t

```

We model absolute population size, and assume that fish age 1 and older have the same survival rate, which varies annually within user-specified lower and upper bounds. Recruitment\index{recruitment} (surviving age-0 fish from the prior year) varies annually about an arbitrary mean according to a lognormal distribution (Section \@ref(Recruit-Fit)). We require a starting population in year 1 and draw the adult survivors from a Poisson distribution, using the same mean as recruitment. Population size in subsequent years is

  $N.a_y = N.r_y + Binomial(N.a_{y-1}, AnnualS_{y-1})$

i.e., incoming recruitment plus surviving adults.

Model parameters are estimated using simulated data from two field studies that are assumed to be independent and would provide information on abundance and survival. The first data source is a two-sample (closed) capture-recapture estimate, assumed to be done at the start of each year. The capture probability, used to generate the two binomally-distributed samples, is set at a relatively high default value to obtain better recruitment estimates. The number of recaptures is also binomally distributed, using the observed marked fraction (on average equal to the capture probability). The second data source is a telemetry study that provides annual information on the number of surviving telemetered fish. For example, fixed receivers might be used to detect migratory fish that return annually to a spawning area. We arbitrarily assume that 20 telemetered fish are released each year. For simplicity, transmitters are assumed to last for the duration of the study. Examine the TelTags matrix to see how the annual releases maintain a good sample size.

Parameter estimates are obtained using uninformative prior distributions for recruitment and annual survival. Population size in year 1 is Poisson distributed, using the capture-recapture point estimate as the expected value. Population size in subsequent years is also drawn from a Poisson distribution, using the population projection equation as the expected value. Both likelihoods use a binomial  distribution\index{Distribution, Binomial} (Section \@ref(BinomialDist)). The probability for the capture-recapture data is the estimated fraction of the population that is marked. The probability for the telemetry data is the estimated annual survival rate.

Following the JAGS section is some R code to compare the true values with model estimates. The <code>par(xpd=TRUE)</code> turns off clipping so that a legend can be added outside the plot boundary (using inset to set the relative coordinates of the legend). For the third plot, the Chapman modification of the capture-recapture estimator [@ricker1975] was used to avoid getting a plotting error due to an infinite point estimate in cases with $m_2=0$. The capture-recapture point estimates (red line) tend to be more variable than the population estimates obtained from the integrated model (black line).

```{r eval=FALSE}
# Load necessary library
library(rjags)
library(R2jags)

# JAGS code ##################################
sink("IPM.txt")
cat("
model {

# Priors

 Mean_lnR.est ~ dnorm(0, 1E-6)  # Recruitment uses ln-scale parameter
 SD_rec.est ~ dunif(0, 10)
 tau <- pow(SD_rec.est, -2)
 for(y in 1:(Y-1)) {
    N.r.est[y] ~ dlnorm(Mean_lnR.est,tau)
     # Hierarchical structure for recruitment
  } #y

for (y in 1:(Y-1)){
  AnnualS.est[y] ~ dunif(0,1)  # Uninformative prior, adult survival
}#y

# Population projection
 N.a.est[1] ~ dpois((n1[1]+1)*(n2[1]+1)/(m2[1]+1)-1) 
  # Year-1 total, Chapman mod in case m2=0
  # Year-1 estimate needed to start model because no prior year information
 for(y in 2:Y) {
     N.a.est[y] ~ dpois(N.r.est[y-1] + (AnnualS.est[y-1]*N.a.est[y-1]))
     # Index shifted by one year for recruitment estimates
     } #y

#Likelihoods
# Two-sample capture-recapture estimate for total pop size at start of year
  for (y in 2:Y){  # m2[1] used above to get starting pop size
    m2[y] ~ dbin((n1[y]/N.a.est[y]), n2[y])
    } #y

# Telemetry information on survival rate
for (y in 1:(Y-1)){
  for (t in ((y+1):Y)){
    TelTags[y,t] ~ dbin(AnnualS.est[t-1],TelTags[y,(t-1)])
  }} # y, t
}
    ",fill=TRUE)
sink()

# Bundle data
jags.data <- list("Y", "n1", "n2", "m2", "TelTags")

# Initial values
jags.inits <- function(){ list(N.r.est= rlnorm(n=(Y-1), meanlog=10, sdlog=1),
                               AnnualS.est=runif(n=(Y-1), min=0, max=1))}

model.file <- 'IPM.txt'

# Parameters monitored
jags.params <- c("AnnualS.est", "N.a.est", "N.r.est", "SD_rec.est")

# Call JAGS from R
jagsfit <- jags(data=jags.data, jags.params, inits=jags.inits,
                n.chains = 3, n.thin=1, n.iter = 40000,
                model.file)
print(jagsfit)
plot(jagsfit)

# Code for plots comparing true values and model estimates
par(mfrow = c(1, 3))
plot(seq(1:(Y-1)), 
     jagsfit$BUGSoutput$mean$AnnualS.est, ylab="Annual S", xlab="Year",
     type="b", pch=19, ylim=c(min(AnnualS,
                                  jagsfit$BUGSoutput$mean$AnnualS.est),
                              max(AnnualS,jagsfit$BUGSoutput$mean$AnnualS.est)))
points(AnnualS, col="blue", type="b", pch=17)

#hist(jagsfit$BUGSoutput$sims.list$MeanR.est, main="", xlab="Mean recruitment")
#abline(v=MeanR, col="red")

plot(seq(1:(Y-1)), 
     jagsfit$BUGSoutput$mean$N.r.est, ylab="Annual recruitment (years 2-Y)", xlab="",
     type="b", pch=19, ylim=c(min(N.r,
                                  jagsfit$BUGSoutput$mean$N.r.est),
                              max(N.r,jagsfit$BUGSoutput$mean$N.r.est)))
points(N.r[2:Y], col="blue", type="b", pch=17)
par(xpd=TRUE) # Turn off clipping to put legend above plot
legend("top",
       legend = c("Est", "True"), col = c("black", "blue"),
       pch = c(19,17), text.col = "black",  horiz = T , inset = c(0, -0.2))
par(xpd=FALSE)

MR_N.hat <- (n1+1)*(n2+1)/(m2+1) -1 # Chapman modification in case m2=0
plot(seq(1:Y), jagsfit$BUGSoutput$mean$N.a.est, 
     ylab="Pop size (includes new recruits)", xlab="Year",
     type="b", pch=19, 
     ylim=c(min(N.a,jagsfit$BUGSoutput$mean$N.a.est,MR_N.hat),
            max(N.a,jagsfit$BUGSoutput$mean$N.a.est,MR_N.hat)))
points(N.a, col="blue", type="b", pch=17) # True adult pop size
points(MR_N.hat, col='red', type="b") # Capture-recapture point estimates
```

Results for the default case show that total population size and annual survival tend to be reliably estimated (Figure \@ref(fig:TwoAgePlot)). Annual recruitment tends to be less well estimated. There is no direct information on recruitment, so the model makes indirect inferences from the total year-y abundance (new age-1 recruits plus 2+ survivors) and telemetry information on survival.

```{r TwoAgePlot, echo=FALSE,warning=FALSE,message=FALSE,error=FALSE, results='hide', fig.cap='Estimates of annual survival (left), recruitment (center), and total population size (right) from a two-age integrated population model fitted to capture-recapture and telemetry data. True values are shown in blue; capture-recapture point estimates (right) are in red.', out.width="90%"}

rm(list=ls()) # Clear Environment
set.seed(11111) # Ensure that book will have a reasonable result
par(mar = c(4, 4, 1, .1))

TwoAge <- function() {
# Priors

 Mean_lnR.est ~ dnorm(0, 1E-6)
   # Recruitment uses ln-scale parameter
 SD_rec.est ~ dunif(0, 10)
 tau <- pow(SD_rec.est, -2)
 for(y in 1:(Y-1)) {
    N.r.est[y] ~ dlnorm(Mean_lnR.est,tau)
     # Hierarchical structure for recruitment
  } #y

for (y in 1:(Y-1)){
  AnnualS.est[y] ~ dunif(0,1)  # Uninformative prior, adult survival
}#y

# Population projection
 N.a.est[1] ~ dpois((n1[1]+1)*(n2[1]+1)/(m2[1]+1)-1) 
  # Year-1 total, Chapman mod in case m2=0
  # Year-1 estimate needed to start model because no prior year information
 for(y in 2:Y) {
     N.a.est[y] ~ dpois(N.r.est[y-1] + (AnnualS.est[y-1]*N.a.est[y-1]))
     # Index shifted by one year for recruitment estimates
     } #y

#Likelihoods
# Two-sample capture-recapture estimate for total pop size at start of year
  for (y in 2:Y){  # m2[1] used above to get starting pop size
    m2[y] ~ dbin((n1[y]/N.a.est[y]), n2[y])
    } #y

# Telemetry information on survival rate
for (y in 1:(Y-1)){
  for (t in ((y+1):Y)){
    TelTags[y,t] ~ dbin(AnnualS.est[t-1],TelTags[y,(t-1)])
  }} # y, t
}

# Simulation to create population size and field data
Y <- 10

LowS <- 0.5 # Adult survival, arbitrary lower bound
HighS <- 0.9 # Arbitrary upper bound
AnnualS <- runif(n=(Y-1), min=LowS, max=HighS)

#MeanR <- 400 # Mean annual recruits
#N.r <- rpois(n=Y, lambda=MeanR)
#N.a <- array(data=NA, dim=Y)
#N.a[1] <- N.r[1] + rpois(n=1, lambda=MeanR) # Arbitrary year-1 population size
#for (y in 2:Y){
#  N.a[y] <- N.r[y] + rbinom(n=1, size=N.a[y-1], prob=AnnualS[y-1])
#} #y

MeanR <- 400 # Arbitrary mean annual recruitment
N.r <- array(data=NA, dim=Y)
SD_rec <- 0.25 # Low level of lognormal error in recruitment
N.r <- trunc(MeanR * rlnorm(n=Y, 0, SD_rec))
N.a <- array(data=NA, dim=Y)
N.a[1] <- N.r[1] + rpois(n=1, lambda=MeanR) # Arbitrary year-1 population size
for (y in 2:Y){
  N.a[y] <- N.r[y] + rbinom(n=1, size=N.a[y-1], prob=AnnualS[y-1])
} #y

# Short-term (closed) two-sample population estimate
CapProb <- 0.2 # Fraction of population sampled
n1 <- rbinom(n=Y, size=N.a, prob=CapProb) # Caught and marked in first sample
n2 <- rbinom(n=Y, size=N.a, prob=CapProb) 
# Caught and examined for marks in second sample
m2 <- rbinom(n=Y, size=n2, prob=n1/N.a) # Marked fish in second sample

# Telemetry information on annual survival
TelRelease <- 20 # Number of telemetry tags to release annually
TelTags <- array(data=NA, dim=c((Y-1),Y))
for (y in 1:(Y-1)){
  TelTags[y,y] <- TelRelease
  for (t in ((y+1):Y)){
    TelTags[y,t] <- rbinom(n=1, size=TelTags[y,(t-1)], prob=AnnualS[t-1])
  }} # y,t

# Load necessary library packages
library(rjags)   # Package for fitting JAGS models from within R
library(R2jags)  # Package for fitting JAGS models. Requires rjags

# JAGS code

# Bundle data
jags.data <- list("Y", "n1", "n2", "m2", "TelTags")

# Initial values.
jags.inits <- function(){ list(N.r.est= rlnorm(n=(Y-1), meanlog=10, sdlog=1),
                               AnnualS.est=runif(n=(Y-1), min=0, max=1))}

# Fit the model
TwoAge_jags <- 
  jags(
    data = jags.data, inits=jags.inits,
    n.chains = 3, n.thin = 1, n.iter = 40000,
    model.file = TwoAge,
    parameters.to.save = c("AnnualS.est", "N.a.est", "N.r.est", "SD_rec.est")
  )

# Code for plots comparing true values and model estimates
par(mfrow = c(1, 3))
plot(seq(1:(Y-1)), 
     TwoAge_jags$BUGSoutput$mean$AnnualS.est, ylab="Annual S", xlab="Year",
     type="b", pch=19, ylim=c(min(AnnualS,
                                  TwoAge_jags$BUGSoutput$mean$AnnualS.est),
                              max(AnnualS,TwoAge_jags$BUGSoutput$mean$AnnualS.est)))
points(AnnualS, col="blue", type="b", pch=17)

plot(seq(1:(Y-1)), 
     TwoAge_jags$BUGSoutput$mean$N.r.est, ylab="Annual recruitment (years 2-Y)", xlab="",
     type="b", pch=19, ylim=c(min(N.r,
                                  TwoAge_jags$BUGSoutput$mean$N.r.est),
                              max(N.r,TwoAge_jags$BUGSoutput$mean$N.r.est)))
points(N.r[2:Y], col="blue", type="b", pch=17)

MR_N.hat <- (n1+1)*(n2+1)/(m2+1) -1 # Chapman modification in case m2=0
plot(seq(1:Y), TwoAge_jags$BUGSoutput$mean$N.a.est, 
     ylab="Pop size (includes new recruits)", xlab="Year",
     type="b", pch=19, 
     ylim=c(min(N.a,TwoAge_jags$BUGSoutput$mean$N.a.est,MR_N.hat),
            max(N.a,TwoAge_jags$BUGSoutput$mean$N.a.est,MR_N.hat)))
points(N.a, col="blue", type="b", pch=17) # True adult pop size
points(MR_N.hat, col='red', type="b") # Capture-recapture point estimates
```

After doing multiple runs with the default settings, try varying the capture probability for the capture-recapture study and the annual release of fish with transmitters. As expected, a higher capture probability improves the estimates of total population size, and more transmitters improve the annual survival estimates.

## Full age-structured model {#CAGEAN}

Another type of IPM is a fish stock assessment\index{stock assessment} that typically uses catch by age and year to estimate population size and fishing mortality rate [@schaub.etal_2024]. Fish stock assessment models share many characteristics with IPMs developed for terrestrial species, but typically have a much greater spatial scale and are more focused on harvst regulation [@schaub.etal_2024]. These models serve as the foundation for much of marine fishery management, as well as in freshwater fisheries where sufficient data are available [e.g., @fielder.bence_2014]. Catch data are a valuable source of information because they reflect absolute abundance. Summing the catches from a cohort (year class) across years (along a diagonal in the catch-at-age matrix) provides a conservative estimate of the initial abundance of that cohort. Early versions of catch-at-age analysis\index{catch-at-age analysis} (virtual population analysis\index{virtual population analysis}; cohort analysis\index{cohort analysis}) were simple upward adjustments of the summed catches in order to account for natural mortality [@ricker1975]. Current catch-at-age analyses have a statistical foundation using either maximum-likelihood or Bayesian estimation [@maunder.punt_2013]. Catch data alone are insufficient to obtain reliable results, so current assessments integrate fishery catch data with axuiliary information from the fishery (e.g., annual level of fishing effort) or surveys [@deriso.etal_1985; @maunder.punt_2013; @schaub.etal_2024]. Most current stock assessments are done using flexible and powerful off-the-shelf software such as AD Model Builder or Stock Synthesis [@fournier.etal_2012; @methot.wetzell_2013].

Here we use JAGS to fit a relatively simple age-structured stock assessment model. We assume a single fishery with gear selectivity that is constant over time. Our auxiliary data are survey catches by age and year, with a survey selectivity pattern that is also constant over time. The default selectivity pattern for the survey provides information mostly on younger ages whereas the fishery primarily exploits older fish. Thus a combined analysis is able to make better inferences than either data set alone. Essentially the goal of the analysis is to find the population estimates that are most consistent with the observed catches and auxiliary data.

The simulation code builds on that used in the cohort-specific catch-curve analysis (\@ref(Cohort)) and recruitment (\@ref(Recruit)) sections. The number of ages is arbitrary but should be sufficient to characterize the selectivity patterns for the fishery and survey.  Annual fishing mortality\index{mortality, fishing} varies randomly between user-set lower and upper bounds. Natural mortality\index{mortality, natural} can be difficult to estimate internally [@quinn.deriso_1999; @schaub.etal_2024], so here it is fixed in the simulation and the fitted model [e.g., using external information such as longevity: @lorenzen1996; @then.etal_2015].

```{r eval=FALSE}
rm(list=ls()) # Clear Environment

# Simulation to create survey and fishery catch-at-age matrices

A <- 8 # Arbitrary number of ages and years (but enough to fit model)
Y <- 12

LowF <- 0.1 # Arbitrary lower bound
HighF <- 0.5 # Arbitrary upper bound

AnnualF <- runif(Y, min=LowF, max=HighF)
AnnualM <- 0.2 # Fixed

# F by age, using separable model
Fishery_k <- 2 # Slope for logistic function
Fishery_a_50 <- 4 # Age at 0.5 selectivity
SelF <- array(data=NA, dim=A) # Fishery selectivity pattern
F_a <- array(data=NA, dim=c(A,Y)) # Matrix for F by age and year
Z_a <- array(data=NA, dim=c(A,Y)) # Matrix for Z by age and year
Exp_C_f <- array(data=NA, dim=c(A,Y)) # Fishery catch-at-age matrix
C_f <- array(data=NA, dim=c(A,Y)) # Obs fishery catch-at-age matrix
Exp_C_s <- array(data=NA, dim=c(A,Y)) # Survey catch-at-age matrix
C_s <- array(data=NA, dim=c(A,Y)) # Obs survey catch-at-age matrix
S_a <- array(data=NA, dim=c(A,(Y-1))) # Matrix for survival rate
for (a in 1:A){
  SelF[a] <- 1/(1+exp(-Fishery_k*(a-Fishery_a_50)))
} #a
#plot(SelF)
for (y in 1:(Y)) {
  for (a in 1:A){
    F_a[a,y] <- AnnualF[y] * SelF[a]
  } #a
} #y
Z_a <- F_a + AnnualM
S_a <- exp(-Z_a)

MeanR <- 1E5 # Arbitrary mean initial cohort size
N <- array(data=NA, dim=c(A, Y))
SD_rec <- 0.25 # Low level of lognormal error in recruitment
N[1,] <- trunc(MeanR * rlnorm(n=Y, 0, SD_rec)) 
# Starting size of each cohort at first age
for (a in 2:A){
  N[a,1] <- trunc(prod(S_a[1:(a-1),1])*MeanR * rlnorm(n=1, 0, SD_rec))
  # Use year-1 S_a to generate year-1 N
  for (y in 2:Y){
    N[a,y] <- rbinom(1, N[(a-1), (y-1)], S_a[(a-1),(y-1)])
  } #y
} #a

for (y in 1:(Y-1)){ # Deaths due to fishing (binomial random variate)
  for (a in 1:(A-1)){
    Exp_C_f[a,y] <- rbinom(1, N[a,y]-N[(a+1),(y+1)],(F_a[a,y]/Z_a[a,y]))
  } #y
} #a
for (y in 1:(Y-1)){ # Random catch last age
  Exp_C_f[A,y] <- rbinom(1,N[A,y],F_a[A,y]*(1-exp(-Z_a[A,y]))/Z_a[A,y])
} #y
for (a in 1:A){ # Random catch last year
  Exp_C_f[a,Y] <- rbinom(1,N[a,Y],F_a[a,Y]*(1-exp(-Z_a[a,Y]))/Z_a[a,Y])
} #y
for (y in 1:Y){
  C_f[,y] <- as.vector(t(rmultinom(n=1, size=sum(Exp_C_f[,y]), 
                                   prob=Exp_C_f[,y])))
} #y

# Generate survey catch-at-age matrix
Survey_q <- 1E-3 # Catchability coefficient (Adjust relative to MeanR)
Survey_k <- -2 # Slope for logistic function
Survey_a_50 <- 4 # Age at 0.5 selectivity
SelS <- array(data=NA, dim=A) # Survey selectivity pattern
for (a in 1:A){
  SelS[a] <- 1/(1+exp(-Survey_k*(a-Survey_a_50)))
} #a
#plot(SelS)

for (y in 1:Y) {
  for (a in 1:A){
    Exp_C_s[a,y] <- Survey_q*SelS[a]*N[a,y]
  } #a
} #y 

for (y in 1:Y){
  C_s[,y] <- as.vector(t(rmultinom(n=1, size=sum(Exp_C_s[,y]), 
                                   prob=Exp_C_s[,y])))
  } #y

```

Survival across ages and years varies randomly according to a binomial distribution\index{distribution, binomial} (Section \@ref(BinomialDist)). Age- and year-specific catch also varies randomly according to a binomial distribution, based on the fraction of deaths due to fishing (F*A/Z). Sampling variation in the fishery and survey catch matrices is based on a multinomial distribution \index{distribution, multinomial} (Section \@ref(Multinomial)). Compare the fishery and survey catch matrices to see the effects of gear selectivity. Also examine the population matrix to see strong and weak year classes moving through the population.

```{r eval=FALSE}
# Load necessary library
library(rjags)
library(R2jags)

# JAGS code ##################################
sink("CatchAge.txt")
cat("
model {
# Model parameters: AnnualF[1:Y], k and a_50 for fishery and survey, survey q,
# Mean recruitment, SD, and annual deviations years 1:Y,
# initial numbers ages 2:A for year 1

 Mean_lnR.est ~ dnorm(0, 1E-6)
   # Recruitment uses ln-scale parameter
 MeanR.est <- exp(Mean_lnR.est)
 SD_rec.est ~ dunif(0, 10)
 tau <- pow(SD_rec.est, -2)
 for(y in 1:Y) {
    Recruit.est[y] ~ dlnorm(Mean_lnR.est,tau)
     # Hierarchical structure for recruitment
    N.est[1,y]<-Recruit.est[y] }

	for(a in 2:A){
	  Initial.est[a-1] ~ dlnorm(0, 1E-6)
	    # Offset index to run Initials vector from 1 to A-1
    N.est[a,1]<-Initial.est[a-1] }

Fishery_k.est ~ dnorm(0, 1E-3)
  # Allow for positive or negative slope for logistic function
Fishery_a_50.est ~ dunif(0,A)
for (a in 1:A){
  SelF.est[a] <- 1/(1+exp(-Fishery_k.est*(a-Fishery_a_50.est)))
  }#a

for (y in 1:Y){
  AnnualF.est[y] ~ dunif(0,2) # Arbitrary range
  for (a in 1:A){
    F_a.est[a,y] <- AnnualF.est[y] * SelF.est[a]
      # Separable model (F[a,y]=Sel[a]*F[y])
    Z_a.est[a,y] <- F_a.est[a,y] + AnnualM
      # Assume that M is known (fixed externally)
  }#a
}#y

Survey_q.est ~ dgamma(1,1) # Uninformative nonnegative prior
Survey_k.est ~ dnorm(0, 1E-3)
  # Allow for positive or negative slope for logistic function
Survey_a_50.est ~ dunif(0, A)
for (a in 1:A){
  SelS.est[a] <- 1/(1+exp(-Survey_k.est*(a-Survey_a_50.est)))
  } #a

	# Population projection
	for(a in 2:A) {
   for(y in 2:Y) {
     N.est[a,y] ~ dbin(exp(-AnnualM-F_a.est[(a-1),(y-1)]),
                         trunc(N.est[a-1,y-1])) } }

	#Likelihoods
	for (a in 1:A) {
   for (y in 1:Y) {
     C_f.est[a,y] <- N.est[a,y]*F_a.est[a,y]*(1-exp(-Z_a.est[a,y]))
                      /Z_a.est[a,y]
     C_f[a,y] ~ dpois(C_f.est[a,y])
     C_s.est[a,y] <- N.est[a,y]*Survey_q.est*SelS.est[a]
     C_s[a,y] ~ dpois(C_s.est[a,y])
   } } # y, a
   for (y in 1:Y){
    N.est.total[y] <- sum(N.est[,y])
   }

}
    ",fill=TRUE)
sink()

# Bundle data
jags.data <- list("Y", "A", "C_f", "C_s", "AnnualM")

# Initial values
jags.inits <- function(){ list(Mean_lnR.est=rlnorm(n=1, meanlog=10, sdlog=1),
                               Recruit.est=rlnorm(n=Y, meanlog=10, sdlog=1),
                               Initial.est=rlnorm(n=(A-1), meanlog=10, sdlog=1),
                               Survey_q.est=rgamma(n=1, shape=1, scale=1),
                               SD_rec.est=runif(n=1, min=0, max=10),
                               AnnualF.est=runif(n=Y, min=0, max=2))}

model.file <- 'CatchAge.txt'

# Parameters monitored
jags.params <- c("AnnualF.est", "Recruit.est", "Initial.est",
                 "SelF.est", "Survey_q.est", "MeanR.est",
                 "SD_rec.est", "SelS.est", "N.est.total")

# Call JAGS from R
jagsfit <- jags(data=jags.data, jags.params, inits=jags.inits,
                n.chains = 3, n.thin=10, n.iter = 200000,
                model.file)
print(jagsfit)
plot(jagsfit)

par(mfrow = c(1, 2))
plot(seq(1:Y), jagsfit$BUGSoutput$mean$AnnualF.est, ylab="Annual F", xlab="Year",
     type="b", pch=19, ylim=c(min(AnnualF,jagsfit$BUGSoutput$mean$AnnualF.est),
                              max(AnnualF,jagsfit$BUGSoutput$mean$AnnualF.est)))
points(AnnualF, col="blue", type="b")

N.total <- array(data=NA, dim=Y)
for (y in 1:Y){
  N.total[y] <- sum(N[,y])
}
plot(seq(1:Y), jagsfit$BUGSoutput$mean$N.est.total, ylab="Total N", xlab="Year",
     type="b", pch=19, ylim=c(min(N.total,jagsfit$BUGSoutput$mean$N.est.total),
                              max(N.total,jagsfit$BUGSoutput$mean$N.est.total)))
points(N.total[], col="blue", type="b")

plot(seq(1:A), jagsfit$BUGSoutput$mean$SelF.est, ylab="Fishery selectivity",
     xlab="Age", type="b", pch=19, ylim=c(0,1))
points(SelF, col="blue", type="b")

plot(seq(1:A), jagsfit$BUGSoutput$mean$SelS.est, ylab="Survey selectivity",
     xlab="Age", type="b", pch=19, ylim=c(0,1))
points(SelS, col="blue", type="b")

```

Convergence (Rhat < 1.1) for this model requires a very large number of updates (and considerable time). Thinning is used to reduce the storage required. The plots following the JAGS code show that there is typically close agreement between the true and estimated values (Figure \@ref(fig:CAGEANPlot)). It is essential to keep in mind that this analysis is being done with an assumed rate of natural mortality\index{natural mortality}. The results of such an analysis, including credible intervals, are conditional on the assumed value. In this case, natural mortality is fixed at the correct value, so these results show the best case in terms of accuracy and precision.

```{r CAGEANPlot, echo=FALSE,warning=FALSE,message=FALSE,error=FALSE, results='hide', fig.cap='Estimates of fishing mortality (left) and total abundance (right) from a stock assessment model fitted to fishery and survey catch-at-age matrices. True values and model estimates are shown in blue and black respectively. ***CHANGE PRODUCTION RUN TO 200K updates', out.width="90%"}

rm(list=ls()) # Clear Environment
set.seed(12345) # Ensure that book will have a reasonable result
par(mar = c(4, 4, 1, .1))

CAGEAN <- function() {
 Mean_lnR.est ~ dnorm(0, 1E-6)
   # Recruitment uses ln-scale parameter
 MeanR.est <- exp(Mean_lnR.est)
 SD_rec.est ~ dunif(0, 10)
 tau <- pow(SD_rec.est, -2)
 for(y in 1:Y) {
    Recruit.est[y] ~ dlnorm(Mean_lnR.est,tau)
     # Hierarchical structure for recruitment
    N.est[1,y]<-Recruit.est[y] }

    for(a in 2:A){
      Initial.est[a-1] ~ dlnorm(0, 1E-6)
        # Offset index to run Initials vector from 1 to A-1
    N.est[a,1]<-Initial.est[a-1] }

Fishery_k.est ~ dnorm(0, 1E-3)
  # Allow for positive or negative slope for logistic function
Fishery_a_50.est ~ dunif(0,A)
for (a in 1:A){
  SelF.est[a] <- 1/(1+exp(-Fishery_k.est*(a-Fishery_a_50.est)))
  }#a

for (y in 1:Y){
  AnnualF.est[y] ~ dunif(0,2) # Arbitrary range
  for (a in 1:A){
    F_a.est[a,y] <- AnnualF.est[y] * SelF.est[a]
      # Separable model (F[a,y]=Sel[a]*F[y])
    Z_a.est[a,y] <- F_a.est[a,y] + AnnualM
      # Assume that M is known (fixed externally)
  }#a
}#y

Survey_q.est ~ dgamma(1,1) # Uninformative nonnegative prior
Survey_k.est ~ dnorm(0, 1E-3)
  # Allow for positive or negative slope for logistic function
Survey_a_50.est ~ dunif(0, A)
for (a in 1:A){
  SelS.est[a] <- 1/(1+exp(-Survey_k.est*(a-Survey_a_50.est)))
  } #a

    # Population projection
    for(a in 2:A) {
   for(y in 2:Y) {
     N.est[a,y] ~ dbin(exp(-AnnualM-F_a.est[(a-1),(y-1)]),
                         trunc(N.est[a-1,y-1])) } }

    #Likelihoods
    for (a in 1:A) {
   for (y in 1:Y) {
     C_f.est[a,y] <- N.est[a,y]*F_a.est[a,y]*(1-exp(-Z_a.est[a,y]))/Z_a.est[a,y]
     C_f[a,y] ~ dpois(C_f.est[a,y])
     C_s.est[a,y] <- N.est[a,y]*Survey_q.est*SelS.est[a]
     C_s[a,y] ~ dpois(C_s.est[a,y])
   } } # y, a
   for (y in 1:Y){
    N.est.total[y] <- sum(N.est[,y])
   }
}

# Simulation to create survey and fishery catch-at-age matrices

A <- 8 # Arbitrary number of ages and years (but enough to fit model)
Y <- 12

LowF <- 0.1 # Arbitrary lower bound
HighF <- 0.5 # Arbitrary upper bound

AnnualF <- runif(Y, min=LowF, max=HighF)
AnnualM <- 0.2 # Fixed

# F by age, using separable model
Fishery_k <- 2 # Slope for logistic function
Fishery_a_50 <- 4 # Age at 0.5 selectivity
SelF <- array(data=NA, dim=A) # Fishery selectivity pattern
F_a <- array(data=NA, dim=c(A,Y)) # Matrix for F by age and year
Z_a <- array(data=NA, dim=c(A,Y)) # Matrix for Z by age and year
Exp_C_f <- array(data=NA, dim=c(A,Y)) # Fishery catch-at-age matrix
C_f <- array(data=NA, dim=c(A,Y)) # Obs fishery catch-at-age matrix
Exp_C_s <- array(data=NA, dim=c(A,Y)) # Survey catch-at-age matrix
C_s <- array(data=NA, dim=c(A,Y)) # Obs survey catch-at-age matrix
S_a <- array(data=NA, dim=c(A,(Y-1))) # Matrix for survival rate
for (a in 1:A){
  SelF[a] <- 1/(1+exp(-Fishery_k*(a-Fishery_a_50)))
} #a
#plot(SelF)
for (y in 1:(Y)) {
  for (a in 1:A){
    F_a[a,y] <- AnnualF[y] * SelF[a]
  } #a
} #y
Z_a <- F_a + AnnualM
S_a <- exp(-Z_a)

MeanR <- 1E5 # Arbitrary mean initial cohort size
N <- array(data=NA, dim=c(A, Y))
SD_rec <- 0.25 # Low level of lognormal error in recruitment
N[1,] <- trunc(MeanR * rlnorm(n=Y, 0, SD_rec)) 
# Starting size of each cohort at first age
for (a in 2:A){
  N[a,1] <- trunc(prod(S_a[1:(a-1),1])*MeanR * rlnorm(n=1, 0, SD_rec))
  # Use year-1 S_a to generate year-1 N
  for (y in 2:Y){
    N[a,y] <- rbinom(1, N[(a-1), (y-1)], S_a[(a-1),(y-1)])
  } #y
} #a

for (y in 1:(Y-1)){ # Deaths due to fishing (binomial random variate)
  for (a in 1:(A-1)){
    Exp_C_f[a,y] <- rbinom(1, N[a,y]-N[(a+1),(y+1)],(F_a[a,y]/Z_a[a,y]))
  } #y
} #a
for (y in 1:(Y-1)){ # Random catch last age
  Exp_C_f[A,y] <- rbinom(1,N[A,y],F_a[A,y]*(1-exp(-Z_a[A,y]))/Z_a[A,y])
} #y
for (a in 1:A){ # Random catch last year
  Exp_C_f[a,Y] <- rbinom(1,N[a,Y],F_a[a,Y]*(1-exp(-Z_a[a,Y]))/Z_a[a,Y])
} #y
for (y in 1:Y){
  C_f[,y] <- as.vector(t(rmultinom(n=1, size=sum(Exp_C_f[,y]), 
                                   prob=Exp_C_f[,y])))
} #y

# Generate survey catch-at-age matrix
Survey_q <- 1E-3 # Catchability coefficient (Adjust relative to MeanR)
Survey_k <- -2 # Slope for logistic function
Survey_a_50 <- 4 # Age at 0.5 selectivity
SelS <- array(data=NA, dim=A) # Survey selectivity pattern
for (a in 1:A){
  SelS[a] <- 1/(1+exp(-Survey_k*(a-Survey_a_50)))
} #a
#plot(SelS)

for (y in 1:Y) {
  for (a in 1:A){
    Exp_C_s[a,y] <- Survey_q*SelS[a]*N[a,y]
  } #a
} #y 

for (y in 1:Y){
  C_s[,y] <- as.vector(t(rmultinom(n=1, size=sum(Exp_C_s[,y]), 
                                   prob=Exp_C_s[,y])))
  } #y

# Load necessary library packages
library(rjags)   # Package for fitting JAGS models from within R
library(R2jags)  # Package for fitting JAGS models. Requires rjags

# JAGS code

# Bundle data
jags.data <- list("Y", "A", "C_f", "C_s", "AnnualM")

# Initial values.
jags.inits <- function(){ list(Mean_lnR.est=rlnorm(n=1, meanlog=10, sdlog=1),
                               Recruit.est=rlnorm(n=Y, meanlog=10, sdlog=1),
                               Initial.est=rlnorm(n=(A-1), meanlog=10, sdlog=1),
                               Survey_q.est=rgamma(n=1, shape=1, scale=1),
                               SD_rec.est=runif(n=1, min=0, max=10),
                               AnnualF.est=runif(n=Y, min=0, max=2))}

# Fit the model
CAGEAN_jags <- 
  jags(
    data = jags.data, inits=jags.inits,
    n.chains = 3, n.thin = 10, n.iter = 20000,
    model.file = CAGEAN,
    parameters.to.save = c("AnnualF.est", "Recruit.est", "Initial.est",
                 "SelF.est", "Survey_q.est", "MeanR.est",
                 "SD_rec.est", "SelS.est", "N.est.total")
  )

# Code for plots comparing true values and model estimates
par(mfrow = c(1, 2))
plot(seq(1:Y), CAGEAN_jags$BUGSoutput$mean$AnnualF.est, ylab="Annual F", xlab="Year",
     type="b", pch=19, ylim=c(min(AnnualF,CAGEAN_jags$BUGSoutput$mean$AnnualF.est),
                              max(AnnualF,CAGEAN_jags$BUGSoutput$mean$AnnualF.est)))
points(AnnualF, col="blue", type="b")

N.total <- array(data=NA, dim=Y)
for (y in 1:Y){
  N.total[y] <- sum(N[,y])
}
plot(seq(1:Y), CAGEAN_jags$BUGSoutput$mean$N.est.total, ylab="Total N", xlab="Year",
     type="b", pch=19, ylim=c(min(N.total,CAGEAN_jags$BUGSoutput$mean$N.est.total),
                              max(N.total,CAGEAN_jags$BUGSoutput$mean$N.est.total)))
points(N.total[], col="blue", type="b")

```

The results of a catch-at-age analysis can be used to explore factors regulating population dynamics. Recruitment in this simulation was assumed to vary about a fixed mean, but in an analysis of real data, recruitment could be compared to spawning stock size (estimated from each year's population vector) or environmental factors.

When experimenting with alternative simulation settings, approximate estimates can be obtained using fewer updates (e.g., 20,000). Rhat values will typically exceed 1.1 but the estimates should show reasonable agreement with the true values. Final estimates can then be obtained using a sufficiently large value for the number of updates.

## Exercises

1.  For the two-age model (Section \@ref(TwoAge)), compare true and estimated mean recruitment in ten trials using the default capture probability of 0.1 versus 0.2 and 0.3. How well does the model estimate total population size and annual survival for each sampling intensity?

2. For the stock assessment model (Section \@ref(CAGEAN)), set the number of updates at a modest level (e.g., 20,000) and compare the level of convergence when fishing mortality is consistently low (e.g., 0.05-0.15) or high (0.95-1.05). Why is there a difference in how readily convergence is achieved for the two cases?
