--- 
title: "A Bayesian Introduction to Fish Population Analysis"
author: "Joseph E. Hightower"
date: "`r Sys.Date()`"
documentclass: krantz
bibliography:
- book.bib
- packages.bib
biblio-style: apalike
link-citations: yes
colorlinks: yes
lot: yes
lof: yes
site: bookdown::bookdown_site
description: "This book is intended to be a bridge between traditional fisheries analytical
  methods and Bayesian statistics. It is a hands-on introduction to models for estimating
  abundance, survival, growth, recruitment, and population trends."
graphics: yes
---

```{r setup, include=FALSE}
options(
  htmltools.dir.version = FALSE, formatR.indent = 2, width = 55, digits = 4
)

# install the packages needed by this book; you fill out c(), e.g. c('ggplot2', 'dplyr')
lapply(c('xfun'), function(pkg) {
  if (system.file(package = pkg) == '') install.packages(pkg)
})
```

# Preface {-}

This book is based in large part on material I developed while teaching (1991-2014) at NC State University. My hope is that the book will be a bridge between traditional fisheries analytical methods and Bayesian approaches that offer many advantages in ecological modeling. The book might be useful as an upper-level undergraduate or early graduate text, or for a working fisheries biologist interested in a hands-on introduction to Bayesian methods.

The general approach for this book follows that used by @kéry_2010 and @kéry.schaub_2011, in that sample data for each Bayesian analysis are produced by simulation. Analyzing simulated data is helpful for learning because the expected results are known. Simulating a field study also provides the flexibility to vary study parameters such as the number of samples or the within-sample variability. Running the simulation and analysis code multiple times under different settings will aid in understanding the methods and their limitations. This brings up a general point about this book: reading about an analysis is not enough -- it is essential to run the code in order to understand the methods. Exercises included at the end of each chapter should also aid in understanding the strengths and weaknesses of each analysis. A solutions manual is available from the author.

## Software information and conventions {-}

This book has been prepared using the **knitr** [@xie2015] and **bookdown** [@R-bookdown] packages. Package names (e.g., **bookdown**) will be indicated by bold text; typewriter font will indicate inline code (e.g., `x^y`) or a function name (followed by parentheses, e.g., <code>mean()</code>).  My R session information is shown below:

```{r}
xfun::session_info()
```

## Acknowledgments {-}

My first exposure to Bayesian statistical methods was via Matthew Krachey, who took the time as a PhD student to teach a few informal sessions on WinBUGS. I soon came to agree with @kéry_2010 that the BUGS language "frees the modeler in you" and would be a valuable tool for research and especially teaching. I incorporated the BUGS language into my graduate fisheries course and soon into research population models. Other colleagues who provided valuable assistance on statistical methods include Beth Gardner, Ken Pollock, and Brian Reich. I also benefited from some of the many excellent reference books on Bayesian methods for ecology [e.g., @mccarthy2007; @kéry_2010; @kéry.schaub_2011].

Anonymous reviewers of previous drafts made a number of helpful suggestions that I have attempted to accommodate. Any remaining errors in the book are mine.

